# Language basics

## Assignments, arithmetic operations 

There are two types of `R` commands: expressions and assignments.

__Expression__

```{r}
cos(pi)
```

__Assignments and expressions__

```{r}
x <- 1 + 2 # or x = 1 + 2
x
y = 4
x == y
```

Using `;` we can type two commands on the same line before executing them:

```{r}
e <- exp (1); log (e)
```

Some examples of arithmetic and Boolean operators:

```{r}
3*4
12/3
2^3
sqrt (16)
1 == 2
1!= 1
FALSE & TRUE # and
FALSE | TRUE # or
```


## Mode, length and class

In R, everything is an _object_. The _mode_ specifies what an object can contain. The main modes are:

- `numeric`: real numbers
- `character`: character strings
- `logical`: logical values true / false
- `list`: list, collection of objects
- `function`: function

`numeric`,` character`, and `logical` objects are _simple_ objects that can contain data of only one type. On the contrary, `list` mobjects are special objects that can contain other objects.


You can access the mode of an object with the `mode ()` function:

```{r}
age = c(33,28, 33) # The concatenation function c() allows to create vectors
mode(age)
names <- c('Daniel', 'Jehanne', 'Romain')
mode(names)
my.list <- list(Names = names, Age = age)
mode(my.list)
mode(is.integer(pi))
mode
```

Besides the mode, an object also has a _length_, defined as the number of elements it contains:

```{r}
length(age)
length(names)
length(my.list)
```

The _class_ of an object specifies its behavior and therefore its way of interacting with operations and functions. An important example are _data frames_: special lists whose elements all have the same length. The class of a data frame is different from that of generic lists and data frames have an indexing system that does not exist for other lists:

```{r}
class(my.list)
my.data.frame = data.frame(names, age)
mode(my.data.frame)
class(my.data.frame)
my.data.frame[1,2] # to extract the 1st element from the 2nd "column"
# Try the following command:
# my.list[1,2]
```

A special object is the missing value `NA`. By default, its mode is `logical`, however` NA` is neither `TRUE` nor` FALSE`. To test if a value is missing we will use the `is.na ()` function:

```{r}
NA == NA # not what we want
is.na(NA)
is.na(mean(c(1,4, NA)))
```
