# Vectors, matrices and functions

## Vectors

The `R` basic object is the _vector_ (a scalar is considered as a vector of length one). The most used function to create a vector is the concatenation:

```{r}
price <- c(150, 162, 155, 157); price
```

__Indexing__ is done through brackets:

```{r}
price[1] # Unlike in Python, the first index is always 1!!
price[c(1.3)]
price[-(1:2)] # to extract all elements except the 1st and 2nd
```

One can also use a boolean indexing vector, the extracted elements are obviously those corresponding to the `TRUE` values. For example to extract prices greater than 156:

```{r}
price > 156 # the boolean vector
price[price > 156]
```

An alternative is given by the `which ()` function which returns the indices whose elements satisfy a logical condition:

```{r}
which(price > 155)
price[which(price > 156)]
```

You can use the indexing to change an element:

```{r}
price [1] <- 0; price
```

It is possible to give labels to the elements of a vector and extract elements based on them:

```{r}
names(price) 
# NULL is a special object with NULL mode that reads "no container"
names(price) <- c('model.1', 'model.2', 'model.3', 'model.4')
price
price['model.3']
```

In a vector, all the elements must have the same mode:

```{r}
x <- c(1,2, 'a', 'b'); x
mode(x)
```

To generate the vector of the first $n$ integers we use the syntax `1:n`

```{r}
1:10
2:6
```

To generate more general sequences we use the `seq()` function:

```{r}
seq(from = 2, to = 20, by = 2) # or more simply seq(2,20,2)
```

We can create a vector of repeated elements with `rep()`:

```{r}
rep(1, len = 3) # same thing as rep (1,3)
rep(NA, 4)
```


## Matrices

A matrix is a vector with a `dim` attribute of length two. All the elements of a matrix therefore have the same mode. To create a matrix:

```{r}
M <- matrix(2:7, nrow = 2, ncol = 3); M
matrix (2:7, nrow = 2, ncol = 3, byrow = TRUE)
```

By default `matrix ()` fills the new matrix one column after another. Indexing is done through brackets:

```{r}
M[2,] # 2nd line
M[, 3] # 3rd column
M[2.3]
M[3]
M[, -2] # to extract all columns except the 2nd
```

To vertically (resp. horizontally) merge two matrices we use `rbind()` (resp. `cbind()`):

```{r}
cbind (M, -M)
rbind(M, 2 * M)
```

## Operations on numerical vectors and matrices

Element wise operations:

```{r}
v <- c(3,4,1,6)
v + 2
v * 2
v * v
v / 2
v / v
v + v^2
sqrt(M)
M * M
# Try the following command:
# M + v
```

Transpose, multiplication, inverse:

```{r}
t(M)
N <- M[, -3]
N %*% diag(1,2) # row by column product matrix
# diag (1,2) builds the 2x2 diagonal matrix where all the
# diagonal elements are equal to 1, ie the 2x2 identity matrix
solve(N)
solve(N) %*% N # checking if solve(N) is the inverse of N
```

__The transpose of a vector is a row matrix: __

```{r}
V <- t(v)
dim(V)
t(V)
```

Pay attention to the following examples:
```{r}
v %*% t(v) # v is considered a column vector!
t(v) %*% v # ditto
diag(1,4) %*% v # ditto
v %*% v # v is both considered a row-vector and a column-vector
```


## Factors

A factor is a vector used to represent qualitative variables, ie a variable with discrete values. Its values, or categories, are called the `levels` in `R`.

```{r}
city <- c('paris', 'lyon', 'lyon', 'paris', 'nantes')
fact.city <- as.factor(city); fact.city
class(fact.city)
levels(fact.city)
```

A factor has the `numeric` mode. The reason for this counter-intuitive fact is that the elements of a factor are represented as integers corresponding to the lexicographic order of their values:

```{r}
mode(fact.city)
as.numeric(fact.city)
```


## User-defined functions

__Example:__

```{r}
my.function <- function(x, y = 10) {# the default value of y is 10
  z = x-y
  return(z)
}
my.function(2)
my.function(2,4)
my.function(y = 1, x = 4)
```

Any variable defined in a function is _local_ and does not appear in the workspace: try to run

```{r, eval = FALSE}
z
```

## Exercises

1. Let $x$ be a vector with the elements of a sample:
```{r, echo = FALSE}
x <- c(45, 63, 17, 32, 54, 57, 41, 29, 34, 37, 18, 39, 46, 43)
x
```
- Write a code to give 
    * the third element of the sample
    * the four elements of the sample
    * the items strictly greater than 35.
    * all elements except those in positions 3, 9 and 12.
- Replace the first element by a missing value and give the position of all elements less than 30. 
2. Write a function `weighted_average` that takes as inputs two vectors $x=(x_1,\ldots,x_n)$ and $w=(w_1,\ldots,w_n)$ and computes the weighted mean
$$
\frac{1}{\sum_{i=1}^n w_i}\sum_{i=1}^nw_ix_i
$$
