# Lists and data frames

## Lists

Lists are special vectors that can store elements of any mode (including other lists). 

```{r}
age = c(33,28, 33)
names <- c('Daniel', 'Jehanne', 'Romain')
my.list <- list(Names = names, Age = age)
```

Like any other vector, a list is indexed by the `[...]` operator, however, note that the result will be a list containing as unique element the desired item:

```{r}
my.list[1]
mode(my.list[1])
```

To get the desired item directly, we therefore use the `[[...]]` operator or the `$` operator followed by the name of the element (if available):

```{r}
my.list[[1]]
my.list$age
```

The elements of a list can have different lengths:
```{r}
city <- c('paris', 'lyon', 'lyon', 'paris', 'nantes')
my.list$city <- city
my.list
```


## Data frames

The most widely data containers is the data frame, a special list of class `data.frame` in which all elements have the same length. For this reason, a data frame is represented in the form of a two-dimensional array whose columns are its elements. Typically, in a data frame the columns represent the __variables__ and the rows the __observations__. Unlike matrices, the elements of a data frame can have different modes.

```{r}
id <- c('id.453', 'id.452', 'id.455', 'id.459', 'id.458', 'id.456', 'id.450', 'id. 451')
age <- c (19, 45, 67, 53, 17, 30, 27, 35)
smoker <- c (TRUE, FALSE, TRUE, TRUE, FALSE, TRUE, TRUE, TRUE)
sex <- c ('f', 'f', 'h', 'h', 'f', 'h', 'f', 'f')
my.db <- data.frame (Id = id, Age = age, Smoker = smoker, Sex = sex); my.db
dim(my.db); nrow(my.db); ncol(my.db)
names(my.db)
```

A data frame being a list, we can extract a column using the `$` operator preceded by the name of the data frame and followed by the name of the column (or variable), or use the operator `[...]`

```{r}
my.db$Sex # a column of characters is automatically transformed into a factor
my.db[, 2]
my.db$Age[my.db$Smoker == FALSE] # simple example of selection
```

The columns are directly accessible in the workspace (without having to type the name of the data frame and the `$`) after having _attached_ the data frame:

```{r}
attach(my.db)
Age
```

To display only the first six lines:
```{r}
head(my.db)
```

Similarly, `tail()` creates a data frame with the last six columns.


## Importing and exporting data

Importing data is a fundamental step in data analysis. To load the data stored in a file (texte, .csv, Excel, ...) into the workspace (ie into memory), you can use the basic function `read.table()`. The three most important arguments are:

- `file`: name (and path) of the file, in quotes
- `header`: are the elements of the first row the names of the columns?
- `sep`: character separating the columns

`read.table()` returns a data frame:

```{r}
url1 <- 'https://raw.githubusercontent.com/vittorioperduca/Introduction-to-R/master/data/iris.txt'
d1 <- read.table(url1,
                 # the first line contains the name of the variables
                 header = TRUE,
                 # values are separated by ;
                 sep = ';') 

head (d1)

url2 <- 'https://raw.githubusercontent.com/vittorioperduca/Introduction-to-R/master/data/heart.txt'
d2 <- read.table(url2,
                 header = TRUE,
                 # variables are separated by a tabulation
                 sep = '\t') 
dim(d2); names(d2)
```

For data stored in the `.Rda` or `.Rdata` format, the import is done with `load()` with the argument `file = filename`. For instance download the `Iris.Rda` file at [github.com/vittorioperduca/Introduction-to-R/blob/master/data/Iris.Rda](https://github.com/vittorioperduca/Introduction-to-R/blob/master/data/Iris.Rda) to your working directory and then try the following:

```{r}
iris_path <- 'data/Iris.Rda' # replace with the file path
load(iris_path)
```


If you want to load `.Rda` or `.Rdata` files directly from an url, don't forget to use the `url()` function (this was not necessary in `read.table()`).

Data can be exported either to a text file (or .csv, Excel ...) using `write.file()` or to `.rda` and `.Rdata` files at using `save()`. In both cases, the two main arguments are 

- `x =` data to save 
- `file =` the name of the file (in quotes).

If the dataset is stored (or must be saved up) locally, it is necessary to know (and be able to modify) the working directory:

```{r}
# getwd() # try on your machine!
# setwd('~/Documents') # to move to the Documents directory
```

Remember that in Linux and macOS machines, `~/` is a shortcut for `/Users/username`. For Windows machines, the address syntax is slightly different. For example we use `\` instead of `/`.

## Exercise

Download the text file [raw.githubusercontent.com/vittorioperduca/Introduction-to-R/master/data/hepatitis.txt](https://raw.githubusercontent.com/vittorioperduca/Introduction-to-R/master/data/hepatitis.txt) to your working directory.

1. Import the dataset in `R`. Warning: missing data were coded with a `?`, read the documentation of `read.table ()`.
2. Find the number of observations, display the names of the variables and the first six observations. Check that the value of `STEROID` for the fourth observation is missing using the appropriate function.
3. Calculate the mean value of `ALBUMIN` in women and men.
4. Create a variable `NSYMP` counting the number of times a variable is equal to` yes` between `FATIGUE` and` MALAISE`. Pay attention to the format of these two variables!
