# Base graphics 

## Basic constructions and scatter plots

We have already seen how to add a line to a scatter plot, to complete the graph we can specify a title, change the axis labels, and add a legend:

```{r}
plot(wt, mpg,
     main = 'Weight and consumption of cars',
     xlab = 'weight', ylab = 'consumption') # xlab = label x axis
abline(h = mean (mpg), col = 2, lty = 2) # h = horizontal line, col = color, lty = line type
legend(x = 'topright', lty = 2, col = 2, legend = 'average consumption')
# x = legend location
```

We can also add a third dimension by coloring the points according to a condition:

```{r}
plot(wt, mpg,
     pch = 19, # pch = type of point
     col = (mpg >= mean (mpg)) + 1 # 1 = black, 2 = red
)
abline(h = mean(mpg), col = 2, lty = 2) # h = horizontal line, col = color, lty = line type
legend(x = 'topright',
       pch = 19,
       col = 2: 1,
       legend = c ('cons> = average', 'cons <average'), # legend = vector with labels
       bty = 'n' # no frame around the legend
)
```

In general, `abline(a, b)` allows to add a line with intercept `a` and slope` b` to an existing plot. For example, an alternative way to add the regression line is as follows:

```{r, eval = FALSE}
intercept = coef(m)[1]
slope = coef(m)[2]
plot(wt, mpg, main = 'Line of least squares')
abline(a = intercept, b = slope)
```

To add any line (not necessarily a straight line) we use the `lines()` function. For example, to add the _lowess_ (locally-weighted polynomial regression) smoothing curve:

```{r}
plot(wt, mpg, main = 'Smoothing curve')
lines(lowess (mpg ~ wt), # lines (): to add a curve
      col = 'blue', # same as col = 4
      lwd = 2) # lwd = line thickness
```

We can add new points using `points()`:

```{r}
plot(wt, mpg, main = 'Smoothing curves')
lines(lowess (mpg ~ wt), col = 'blue', lwd = 2)
outliers = cbind(c(4.5,4.6,5), c(26,26,30))
points(outliers, pch = 2) # pch = 2 for triangles
lines(lowess (c (wt, outliers[, 1]),
              c (mpg, outliers[, 2])),
      col = 2, lwd = 2)
legend (x = 'bottomleft', col = c (4,2), lwd = 2, legend = c ('without outliers', 'with outliers'), bty = 'n')
```

We can control the range of axes using `xlim()` and `ylim()`:

```{r}
plot(wt, mpg, main = 'Same data with different scale',
     xlim = c (0, 25),
     ylim = c (0, 100))
```


## Line graphs

The `lines()` function adds a line to an existing graph and cannot be used to create a new graph. To draw a line connecting the successive elements of a vector we will rather use `plot(x = vector, type = l)`:

```{r}
mydata = cos(2 * pi / 12 * (1:36)) + rnorm (36,0, .2) # data simulation
plot(mydata, type = 'l', main = 'type = l')
```

The `type` parameter allows you to connect (or not) the points in different ways:

```{r}
par(mfrow = c (2,2)) # explained in the next section
plot(mydata, type = 'p', main = 'type = p') # default
plot(mydata, type = 'o', main = 'type = o')
plot(1: 10, type = 'S', main = 'type = S') # steps 1
plot(1: 10, type = 's', main = 'type = s') # steps 2
```

With `plot = 'h'` we get a bar plot:

```{r}
plot (1: 10, type = 'h', main = 'type = h')
```


## Graphical parameters

We have already seen that it is possible to control certain graphical elements using parameters:

Element | Parameter
------- | ---------
point | `pch`
type of line graph | `type`
color | `col`
line | `lty`,` lwd`
axis label | `xlab`,` ylab`
axis dimensions | `xlim`,` ylim`
label dimensions | `cex`
orientation axes labels | `las`

See `?par` for a description of the values that these parameters can take.

To impose parameters on all the graphics produced during a session, we will use the `par()` function. `par()` is often used to view two or more plots in the same window with the parameter `mfrow = c(l, c)`. In this case the graphs are displayed in a grid with `l` rows and `c` columns. Try the following:

```{r}
# opar = par () # to be able to reset with the initial parameters
par(mfrow = c (1,2))
plot(wt, mpg, main = 'Weight, consumption')
boxplot(hp, main = 'Power')
```

Changes are implemented until when the session is closed, or the graphics engine is reset by `dev.off()` or by clicking on the `Clear all plots` brush in `RStudio`.

`par()` is also used to change the size of the margins. This is sometimes useful when the labels on the axes do not fit in the window, as in this example:

```{r}
mydata = as.factor(sample (x = c ('Paris', 'London', 'Brussels', 'Washington'),
                           replace = T,
                           prob = c (.1, .5, .2, .1),
                           size = 100)) # simulation of a qualitative variable
barplot(summary (mydata),
        width = .1,
        horiz = T,
        las = 2)
```

We modify the parameter which sets the left margin:

```{r}
par(mar = c(5, 6, 4, 2) + 0.1) # the second component of mar gives the left margin
# Default is c (5, 4, 4, 2) + 0.1
barplot(summary (mydata),
        width = .1,
        horiz = T,
        las = 2)
```

## Histograms

As we have already noted in the subsection \@ref(quant-var), by default, `hist()` displays the histogram with the counts. To display the histogram in the density scale we use the option `freq = FALSE`. In this case, the area of each rectangle will be equal to the proportion of observations in the corresponding class (so that the total area of all the rectangles is one).

```{r, eval = FALSE}
par(mfrow = c (1,2))
hist(mpg) # freq = TRUE
hist(mpg, freq = FALSE)
```
     
To superimpose the curve of a given density:

```{r}
mydata = rnorm (1000,0,1) # simulation of 1000 observations N (0,1)
hist(mydata, freq = F, main = 'Histogram and theoretical density')
curve(dnorm, # the density function of N (0,1)
      from = -5, to = 5, # the range
      add = TRUE) # because we want to add the density of N (0,1) to the histogram
```

Rather than viewing the histogram of the data, we can show the estimated density (using kernel density estimation methods):

```{r}
hist(mpg, freq = F, main = 'Histogram and estimated density')
lines(density (mpg), col = 2)
```